---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data


* * *

## Part 2: Research questions

**Research quesion 1:**
My first question is: Is there a relationship between income and number of days of poor (specified as 'not good') health? 

Within the data, there are two categorical variables associated with health; number of days of poor (not good) physical health and number of days of poor (not good) mental health. 

It will be interesting to understand if mental or physical health illness, or both, are more prevalent within certain income types, as well as what the average looks like for survey participants.
**Research quesion 2:**

**Research quesion 3:**


* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**
In the first instance, I want to exclude data that is NA, as this is not relevant to my question, for example, we only want to compare mental and physical health where we know the income amount.
```{r}
brfss2013 %>% filter(!is.na(income2),!is.na(physhlth),!is.na(menthlth)) %>%
 group_by(income2,physhlth,menthlth) %>% summarise(count=n())
```

Next, I am interested to understand what the mean, median and standard deviation looks like for days of not good mental and physical health. This will help me understand if there are any initial skews within the data.

```{r}
brfss2013 %>% filter(!is.na(menthlth)) %>% summarise(mean_mental = mean(menthlth), median_mental = median(menthlth), sd_mental=sd(menthlth), n=n())
```

We can see that there is a big difference between the mean and the median. This suggests a right skew, as the mean is greater than the median. We can confirm this by building a histogram:

```{r}
ggplot(data=subset(brfss2013, !is.na(menthlth)), aes(x=menthlth)) + geom_histogram(binwidth=0.5) + xlim(0, 50)
```

This right skew is confirmed by the histogram. Understanding the skew of the data helps us to see that this is not a normal distribution. We can then see if there are any differences with the physical health variable.

```{r}
brfss2013 %>% filter(!is.na(physhlth)) %>% summarise(mean_physical = mean(physhlth), median_physical = median(physhlth), sd_physical=sd(physhlth), n=n())
```

Again, we can see that the mean is much greater than the median. Creating another histogram, we can then view the distribution of this variable.

```{r}
ggplot(data=subset(brfss2013, !is.na(physhlth)), aes(x=physhlth)) + geom_histogram(binwidth=0.5) + xlim(0, 50)
```


**Research quesion 2:**

```{r}

```



**Research quesion 3:**

```{r}

```

